---
title: "Clean Leading GBM Model: SIPP to ACS"
format: html
---

## 1. Setup

```{r}
# Core packages
library(dplyr)
library(readr)
library(gbm)
library(caret)

# Set project paths
# This lets the same file run on Windows or Mac.
if (Sys.info()["sysname"] == "Windows") {
  base_dir <- "G:/Shared drives/Undocu Research"
} else {
  base_dir <- "~/Library/CloudStorage/GoogleDrive-vsovero@ucr.edu/Shared drives/Undocu Research"
}

data_path <- file.path(base_dir, "Data")
```

## 2. Read cleaned SIPP and ACS files

```{r}
# SIPP donor/training file created from Stata cleaning
sipp <- read_csv(
  file.path(data_path, "(Step 1 output) Core_TM SIPP 2008 Wave 2.csv"),
  show_col_types = FALSE
)

# ACS target sample created from Stata cleaning
acs <- read_csv(
  file.path(data_path, "ACS Target Sample.csv"),
  show_col_types = FALSE
)
```

## 3. Define model variables

```{r}
# These are the predictor variables used in BOTH SIPP and ACS.
# Race enters as separate race dummies, not as the raw race code.
# years_us_missing is included because missing years_us was filled with 0 in Stata.
features <- c(
  "age",
  "fem",
  "married",
  "cit_spouse",
  "medicaid",
  "nonfluent",
  "spanish_hispanic_latino",
  "central_latino",
  "bpl_asia",
  "household_size",
  "poverty",
  "asian",
  "black",
  "white",
  "other_race",
  "employed",
  "years_us",
  "years_us_missing",
  "yrsed"
)

# Explicit model formula.
# This is intentionally written out for transparency.
gbm_formula <- undocu_likely ~
  age +
  fem +
  married +
  cit_spouse +
  medicaid +
  nonfluent +
  spanish_hispanic_latino +
  central_latino +
  bpl_asia +
  household_size +
  poverty +
  asian +
  black +
  white +
  other_race +
  employed +
  years_us +
  years_us_missing +
  yrsed
```

## 4. Prepare SIPP and ACS model samples

```{r}

# Explicit SIPP age restriction.
# The older code implicitly dropped ages 0-14 because yrsed was missing
# outside the education-question universe. This makes that restriction visible.
sipp <- sipp %>%
  filter(age >= 18)

# Keep SIPP rows that are complete on the outcome, model predictors, and college.
# college is not used for training, but is needed later to define the
# target-like SIPP validation group.
sipp <- sipp[complete.cases(sipp[, c("undocu_likely", features, "college")]), ]

# Keep ACS rows that are complete on merge keys and model predictors.
# These are the rows that will receive predicted probabilities.
acs <- acs[complete.cases(acs[, c("year", "serial", "pernum", features)]), ]
```

## 5. Split SIPP into training and test samples

```{r}
set.seed(1)

# Split SIPP into 70% training and 30% test.
# The model is trained only on train_gbm.
# The test sample is used to define probability thresholds.
train_index <- createDataPartition(
  sipp$undocu_likely,
  p = 0.70,
  list = FALSE
)

train_gbm <- sipp[train_index, ]
test_gbm  <- sipp[-train_index, ]

cat("Training N:", nrow(train_gbm), "\n")
cat("Test N:", nrow(test_gbm), "\n")
cat("Training positive share:", mean(train_gbm$undocu_likely), "\n")
cat("Test positive share:", mean(test_gbm$undocu_likely), "\n")
```

## 6. Train full-sample GBM model

```{r}
set.seed(1)

# Train one GBM model on the full SIPP training sample.
# These parameters match the leading GBM setup used previously.
gbm_model <- gbm(
  formula = gbm_formula,
  data = train_gbm,
  distribution = "bernoulli",
  verbose = FALSE,
  cv.folds = 10,
  n.trees = 3500,
  shrinkage = 0.005,
  interaction.depth = 6
)

# Use cross-validation inside gbm to choose the number of trees.
optimal_trees_gbm <- gbm.perf(
  gbm_model,
  method = "cv",
  plot.it = FALSE
)

cat("Optimal number of trees:", optimal_trees_gbm, "\n")

# Variable importance summary
summary(gbm_model)
```

## 7. Predict probabilities in SIPP test sample and ACS target sample

```{r}
# Predicted probability for held-out SIPP test observations.
test_gbm$gbm_undocu_p <- predict(
  gbm_model,
  newdata = test_gbm,
  n.trees = optimal_trees_gbm,
  type = "response"
)

# Predicted probability for ACS target observations.
acs$gbm_undocu_p <- predict(
  gbm_model,
  newdata = acs,
  n.trees = optimal_trees_gbm,
  type = "response"
)
```

## 8. Define quartiles using full SIPP test sample

```{r}
# Full-SIPP quartiles reproduce the original leading-file logic.
# The cutpoints are based on the full held-out SIPP test sample.
# ACS cases are then classified using these same cutpoints.
full_sipp_breaks <- quantile(
  test_gbm$gbm_undocu_p,
  probs = c(0, .25, .50, .75, 1),
  na.rm = TRUE
)

# Extend endpoints so ACS values outside the SIPP test range are not coded as NA.
full_sipp_breaks[1] <- -Inf
full_sipp_breaks[5] <- Inf

test_gbm$gbm_undocu_q_full <- cut(
  test_gbm$gbm_undocu_p,
  breaks = full_sipp_breaks,
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4")
)

acs$gbm_undocu_q_full <- cut(
  acs$gbm_undocu_p,
  breaks = full_sipp_breaks,
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4")
)

# Check how well the full-SIPP quartiles rank true positives in SIPP.
full_sipp_quartiles <- test_gbm %>%
  group_by(gbm_undocu_q_full) %>%
  summarise(
    n = n(),
    undocu_likely_n = sum(undocu_likely == 1),
    undocu_likely_share = mean(undocu_likely == 1),
    mean_p = mean(gbm_undocu_p),
    .groups = "drop"
  )

# Check how ACS cases are distributed across the full-SIPP quartile cutpoints.
acs_full_quartiles <- acs %>%
  group_by(gbm_undocu_q_full) %>%
  summarise(
    n = n(),
    share = n() / nrow(acs),
    mean_p = mean(gbm_undocu_p),
    .groups = "drop"
  )

full_sipp_quartiles
acs_full_quartiles
```

## 9. Define quartiles using target-like SIPP test sample

```{r}
# Target-like SIPP reference group.
# This is NOT used to train the model.
# It is only used to create a second set of quartile cutpoints that better
# matches the ACS target sample: college graduates, employed, ages 22-55.
sipp_targetlike_test <- test_gbm %>%
  filter(
    college == 1,
    employed == 1,
    age >= 22,
    age <= 55
  )

targetlike_breaks <- quantile(
  sipp_targetlike_test$gbm_undocu_p,
  probs = c(0, .25, .50, .75, 1),
  na.rm = TRUE
)

# Extend endpoints so ACS values outside the SIPP target-like range are retained.
targetlike_breaks[1] <- -Inf
targetlike_breaks[5] <- Inf

sipp_targetlike_test$gbm_undocu_q_targetlike <- cut(
  sipp_targetlike_test$gbm_undocu_p,
  breaks = targetlike_breaks,
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4")
)

acs$gbm_undocu_q_targetlike <- cut(
  acs$gbm_undocu_p,
  breaks = targetlike_breaks,
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4")
)

# Check how well the target-like quartiles rank true positives in the
# target-like SIPP test group.
targetlike_sipp_quartiles <- sipp_targetlike_test %>%
  group_by(gbm_undocu_q_targetlike) %>%
  summarise(
    n = n(),
    undocu_likely_n = sum(undocu_likely == 1),
    undocu_likely_share = mean(undocu_likely == 1),
    mean_p = mean(gbm_undocu_p),
    .groups = "drop"
  )

# Check how ACS cases are distributed across the target-like SIPP cutpoints.
acs_targetlike_quartiles <- acs %>%
  group_by(gbm_undocu_q_targetlike) %>%
  summarise(
    n = n(),
    share = n() / nrow(acs),
    mean_p = mean(gbm_undocu_p),
    .groups = "drop"
  )

targetlike_sipp_quartiles
acs_targetlike_quartiles
```

## 10. Define high-recall thresholds

```{r}
# Original high-recall definition.
# Sort true positives in the full SIPP test sample by predicted probability.
# The cutoff is the lowest probability among the top 75% of true positives.
top_75_full <- test_gbm %>%
  filter(undocu_likely == 1) %>%
  arrange(desc(gbm_undocu_p)) %>%
  mutate(
    rank = row_number(),
    cumulative_p = rank / n(),
    top_75 = cumulative_p <= 0.75
  ) %>%
  filter(top_75 == TRUE)

threshold_75_full <- min(top_75_full$gbm_undocu_p)

# Target-like high-recall threshold.
# Same logic, but only among true positives in the target-like SIPP test group.
top_75_targetlike <- sipp_targetlike_test %>%
  filter(undocu_likely == 1) %>%
  arrange(desc(gbm_undocu_p)) %>%
  mutate(
    rank = row_number(),
    cumulative_p = rank / n(),
    top_75 = cumulative_p <= 0.75
  ) %>%
  filter(top_75 == TRUE)

threshold_75_targetlike <- min(top_75_targetlike$gbm_undocu_p)

cat("Full-SIPP high-recall threshold:", threshold_75_full, "\n")
cat("Target-like high-recall threshold:", threshold_75_targetlike, "\n")
```

## 11. Export ACS predictions for Stata

```{r}
# Export file used by the Stata EO_Final reintegration do-file.
#
# Legacy downstream variables:
# gbm_high_prob   = Q4 using full-SIPP test quartiles
# gbm_low_prob    = Q1 using full-SIPP test quartiles
# gbm_high_recall = high-recall threshold from full SIPP test positives
#
# Additional target-like variables are exported separately so they can be used
# for robustness or comparison without replacing the legacy definitions.
acs_gbm_export <- acs %>%
  mutate(
    gbm_high_prob = ifelse(gbm_undocu_q_full == "Q4", 1, 0),
    gbm_low_prob = ifelse(gbm_undocu_q_full == "Q1", 1, 0),
    gbm_high_recall = ifelse(gbm_undocu_p >= threshold_75_full, 1, 0),

    gbm_high_prob_targetlike = ifelse(gbm_undocu_q_targetlike == "Q4", 1, 0),
    gbm_low_prob_targetlike = ifelse(gbm_undocu_q_targetlike == "Q1", 1, 0),
    gbm_high_recall_targetlike = ifelse(gbm_undocu_p >= threshold_75_targetlike, 1, 0)
  ) %>%
  select(
    gbm_undocu_p,
    gbm_undocu_q_full,
    gbm_high_prob,
    gbm_low_prob,
    gbm_high_recall,
    gbm_undocu_q_targetlike,
    gbm_high_prob_targetlike,
    gbm_low_prob_targetlike,
    gbm_high_recall_targetlike,
    year,
    serial,
    pernum
  )

write_csv(
  acs_gbm_export,
  file.path(data_path, "ACS_SIPP_gbm.csv")
)
```

## 12. Save model and threshold metadata

```{r}
# Save the model and threshold choices so later files do not have to re-create them.
model_metadata <- list(
  features = features,
  optimal_trees_gbm = optimal_trees_gbm,
  full_sipp_breaks = full_sipp_breaks,
  targetlike_breaks = targetlike_breaks,
  threshold_75_full = threshold_75_full,
  threshold_75_targetlike = threshold_75_targetlike,
  training_n = nrow(train_gbm),
  test_n = nrow(test_gbm),
  created_at = Sys.time()
)

saveRDS(gbm_model, file.path(data_path, "gbm_model_full_sipp.rds"))
saveRDS(model_metadata, file.path(data_path, "gbm_model_metadata.rds"))
```

```{r}
# Apply full-SIPP quartile cutpoints to the target-like SIPP test sample
sipp_targetlike_test$gbm_undocu_q_full <- cut(
  sipp_targetlike_test$gbm_undocu_p,
  breaks = full_sipp_breaks,
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4")
)

sipp_targetlike_using_full_quartiles <- sipp_targetlike_test %>%
  group_by(gbm_undocu_q_full) %>%
  summarise(
    n = n(),
    share = n() / nrow(sipp_targetlike_test),
    undocu_likely_n = sum(undocu_likely == 1),
    undocu_likely_share = mean(undocu_likely == 1),
    mean_p = mean(gbm_undocu_p),
    .groups = "drop"
  )

sipp_targetlike_using_full_quartiles
```
