---
title: "ML Undocu imputation (SIPP to ACS)"
author: "Mario Arce Acosta"
date: "2025-03-17"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load raw data}
data_path <- "G:/Shared drives/Undocu Research/Data"
figures_path <- "G:/Shared drives/Undocu Research/Output/Figures"

setwd(data_path)

library(arrow)
sipp08_2 <- read_parquet("sipp_2008_w2_merged.parquet")
```

```{r Data preparation}
library(dplyr)

# Define the list of raw variables that need to be numeric
numeric_vars <- c(
  "TIMSTAT", "EADJUST", "EEDUCATE", "TMOVEUS", "RHCALYR", "EMS", 
  "ESEX", "ERACE", "EORIGIN", "TBRSTATE", "EHOWWELL", "TLANG1", 
  "TFIPSST", "EBORNUS", "ECITIZEN", "cit_spouse", "RMESR", 
  "THEARN", "RHPOV", "EHHNUMPP", "RCUTYP57", "RCUTYP58", 
  "ECRMTH", "RCUTYP01", "RCUTYP03", "EAFNOW", "EAFEVER", "TAGE", "TPEARN", "EHRSALL"
)

sipp08_2 <- sipp08_2 %>%
  # STEP 1: Pre-clean edge cases and convert everything to numeric
  mutate(
    # Fix the baby age string before converting to numeric
    TAGE = ifelse(TAGE == "Less than 1 full year old", "0", TAGE),
    # Convert all listed variables to numeric at once
    across(all_of(numeric_vars), as.numeric)
  ) %>%
  
  # STEP 2: The clean logic block (no more as.numeric() needed here!)
  mutate(
    # 1. IMMIGRATION & LEGAL STATUS
    undocu_entry = ifelse(TIMSTAT == 2, 1, 0), 
    undocu_likely = as.factor(ifelse(TIMSTAT == 2 & EADJUST == 2, 1, 0)),
    
    # 2. EDUCATION 
    education = case_when(
      EEDUCATE %in% 31:38 ~ "No HS diploma",          
      EEDUCATE %in% c(39, 41) ~ "HS diploma",         
      EEDUCATE == 40 ~ "Some college",                
      EEDUCATE %in% 42:43 ~ "Associate's",            
      EEDUCATE == 44 ~ "Bachelor's",                  
      EEDUCATE == 45 ~ "Master's",                    
      EEDUCATE %in% 46:47 ~ "PhD/Professional",       
      TRUE ~ "Unknown" 
    ),
    yrsed = case_when(
      EEDUCATE == 36 ~ 10,                            
      EEDUCATE == 37 ~ 11,                            
      EEDUCATE %in% c(38, 39, 40, 41) ~ 12,           
      EEDUCATE == 32 ~ 2.5,                           
      EEDUCATE == 33 ~ 5.5,                           
      EEDUCATE == 34 ~ 7.5,                           
      EEDUCATE == 35 ~ 9,                             
      EEDUCATE == 31 ~ 0,                             
      EEDUCATE %in% 42:43 ~ 14,                       
      EEDUCATE %in% c(44, 46) ~ 16,                   
      EEDUCATE == 45 ~ 17.5,                          
      EEDUCATE == 47 ~ 22,                            
      TRUE ~ NA_real_
    ),
    college = ifelse(EEDUCATE %in% 44:47, 1, 0),
    hs_only = ifelse(EEDUCATE %in% 39:43, 1, 0),
    
    # 3. DEMOGRAPHICS
    # Encoding: All immig_year encoding is an approximation
    immig_yr = case_when(
      is.na(TMOVEUS) ~ 0,
      TMOVEUS == 1~1961, ## Technically
      TMOVEUS == 2~1966,
      TMOVEUS == 3~1971,
      TMOVEUS == 4~1976,
      TMOVEUS == 5~1980,
      TMOVEUS == 6~1982,
      TMOVEUS == 7~1984,
      TMOVEUS == 8~1987,
      TMOVEUS == 9~1989,
      TMOVEUS == 10~1991,
      TMOVEUS == 11~1993,
      TMOVEUS == 12~1995,
      TMOVEUS == 13~1998,
      TMOVEUS == 14~1999,
      TMOVEUS == 15~2000,
      TMOVEUS == 16~2001,
      TMOVEUS == 17~2002,
      TMOVEUS == 18~2004,
      TMOVEUS == 19~2005,
      TMOVEUS == 20~2006,
      TMOVEUS == 21~2007,
      TMOVEUS == 22~2009,
      TRUE ~ 0 # Default case
    ),
    years_us = RHCALYR - immig_yr,
    
    married = ifelse(EMS %in% c(1, 2), 1, 0),
    fem = ifelse(ESEX == 2, 1, 0),
    
    race = case_when(
      ERACE == 3 ~ "Asian",
      ERACE == 2 ~ "Black",
      ERACE == 1 ~ "White",
      ERACE == 4 ~ "Other",
      TRUE ~ "Unknown"
    ),
    asian = ifelse(ERACE == 3, 1, 0),
    black = ifelse(ERACE == 2, 1, 0),
    white = ifelse(ERACE == 1, 1, 0),
    other_race = ifelse(ERACE == 4, 1, 0),
    
    spanish_hispanic_latino = ifelse(EORIGIN == 1, 1, 0),
    central_latino = ifelse(TBRSTATE == 570 & EORIGIN == 1, 1, 0), # Update 570 to your codebook value
    bpl_asia = ifelse(TBRSTATE %in% c(565:567), 1, 0), # Update to your codebook values
    
    # 4. LANGUAGE 
    english_difficult = ifelse(EHOWWELL %in% c(3, 4), 1, 0),
    nonfluent = ifelse(EHOWWELL %in% c(3, 4), 1, 0),
    english_home = ifelse(TLANG1 == -1, 1, 0),
    
    # 5. GEOGRAPHY & CITIZENSHIP
    top_ten_states = ifelse(TFIPSST %in% c(4, 6, 12, 13, 17, 34, 36, 37, 48, 53), 1, 0),
    bpl_usa = ifelse(EBORNUS == 1, 1, 0),
    citizen = ifelse(ECITIZEN == 1, 1, 0),
    cit_spouse = ifelse(cit_spouse == 1, 1, 0),
    
    # 6. EMPLOYMENT & INCOME 
    employed = ifelse(RMESR %in% c(1, 2, 3), 1, 0),
    poverty = ifelse(THEARN < RHPOV, 1, 0),
    household_size = EHHNUMPP,
    
    # --- INFLATION ADJUSTMENT (2008 to 2026 Dollars) ---)
    adj_income = TPEARN * (326.588 / 211.080),
    # Average income per month divided by average hours per week
    ## 4.34524 constant used to convert month to week
    # WAGE & HOUR CLEANING
    # Monthly Income / (Weekly Hours * 4.34524 weeks per month)
    adj_hourly = ifelse(adj_income <= 0 | EHRSALL <= 0, 
                        NA_real_, 
                        adj_income / (EHRSALL * 4.34524)),
    
    ln_adj = log(adj_hourly),
    
    # 7. BENEFITS
    medicaid = ifelse(RCUTYP57 == 1, 1, 0),
    health_ins = ifelse(RCUTYP57 == 1 | RCUTYP58 == 1, 1, 0),
    medicare = ifelse(ECRMTH == 1, 1, 0),
    social_security = ifelse(RCUTYP01 == 1 | RCUTYP03 == 1, 1, 0),
    armed_forces = ifelse(EAFNOW == 1 | EAFEVER == 1, 1, 0)
  ) %>%
  filter(TAGE >= 18 & TAGE <= 55 &  TPEARN >=0)


# Add this AFTER your main mutate block, before you train or predict!
sipp08_2 <- sipp08_2 %>%
  mutate(
    # Explicitly define the levels so empty categories are preserved as 0-counts
    education = factor(education, levels = c("No HS diploma", "HS diploma", 
                                             "Some college", "Associate's", 
                                             "Bachelor's", "Master's", 
                                             "PhD/Professional", "Unknown")),
    
    race = factor(race, levels = c("Asian", "Black", "White", "Other", "Unknown"))
  )

# --- POST-MUTATE LOGICAL CORRECTIONS ---
sipp08_2$bpl_foreign <- ifelse(sipp08_2$bpl_usa == 1, 0, 1)

# Because undocu_likely is a factor, we must replace with the string "0"
sipp08_2$undocu_likely <- replace(sipp08_2$undocu_likely, sipp08_2$immig_yr <= 1961, "0")
sipp08_2$undocu_likely <- replace(sipp08_2$undocu_likely, sipp08_2$armed_forces == 1 | sipp08_2$social_security == 1, "0")

sipp08_2$years_us <- ifelse(sipp08_2$years_us %in% c(2008, 2009, -1), NA, sipp08_2$years_us)
sipp08_2$age <- sipp08_2$TAGE

sipp08_2$undocu_logical <- ifelse(sipp08_2$citizen == 0 & (sipp08_2$armed_forces == 0 | sipp08_2$medicare == 0 | sipp08_2$social_security == 0), 1, 0)
sipp08_2$id <- seq_len(nrow(sipp08_2))

# 3. VARIABLE LISTS
variable_lists <- list(
  base = c("undocu_likely", "undocu_logical", "bpl_foreign", "medicaid", "central_latino", "bpl_asia", "age", "fem", "married", "cit_spouse", "nonfluent", "spanish_hispanic_latino", "household_size", "top_ten_states", "poverty", "asian", "black", "white", "other_race", "employed", "ln_adj", "years_us", "yrsed", "college"),
  
  descriptive = c("undocu_likely", "undocu_logical", "bpl_foreign", "medicaid", 
                  "central_latino", "bpl_asia", "age", "fem", "married", "cit_spouse", 
                  "nonfluent", "spanish_hispanic_latino", "household_size", "poverty", 
                  "asian", "black", "white", "other_race", "employed", "ln_adj", "years_us", "yrsed")
)

# 4. DATASET CREATOR
create_datasets <- function(data, variables) {
  datasets <- list()
  
  # Filter now works perfectly because undocu_logical is a true numeric
  undocu_filter <- data[data$undocu_logical == 1, ]
  
  # Using modern dplyr 'across()' instead of deprecated 'mutate_at'
  datasets$dTable <- as.data.frame(undocu_filter[, variables$descriptive]) %>%
    mutate(across(c(everything(), -undocu_likely), as.numeric)) %>%
    na.omit()
  
  datasets$sample <- as.data.frame(undocu_filter[, variables$base]) %>%
    na.omit()
  
  datasets$knn <- as.data.frame(undocu_filter[, variables$base]) %>%
    mutate(across(c(everything(), -undocu_likely), as.numeric)) %>%
    na.omit()
  
  # Demographic subsets now work flawlessly with == 0 or == 1
  datasets$noncit <- data[data$citizen == 0, ]
  datasets$central_latino <- data[data$central_latino == 1, ]
  datasets$spanish_hispanic_latino <- data[data$spanish_hispanic_latino == 1, ]
  datasets$top_states <- data[data$top_ten_states == 1, ]
  
  return(datasets)
}

# Execute
all_data <- create_datasets(sipp08_2, variable_lists)

all_data_college <- lapply(all_data, function(x) {
  if (is.data.frame(x) && "college" %in% names(x)) {
    x %>% filter(college == 1)
  } else {
    x  
  }
})

### Save cleaned data ###
setwd("G:/Shared drives/Undocu Research/Data")
write_parquet(all_data$sample, "cleaned_SIPP.parquet")
# Save output
# write.csv(sipp08_2, "SIPP08_2.csv", row.names = FALSE)
```

# Logistic

```{r Logistic Regression}
library(stargazer)
library(caret)
library(xtable)
library(ROCR)


levels(all_data$sample$undocu_likely) <- make.names(levels(all_data$sample$undocu_likely))


set.seed(1)
train_index_logistic <- createDataPartition(all_data$sample$undocu_likely, p = 0.7, list = FALSE)
train_logistic <- all_data$sample[train_index_logistic, ]
test_logistic <- all_data$sample[-train_index_logistic, ]

## Create trainControl object
control <- trainControl(
    method = "cv",
    number = 10,  
    summaryFunction = twoClassSummary,
    classProbs = TRUE,
    sampling = "up"
)

## Train glm with custom trainControl
logistic_model <- train(undocu_likely ~ age + fem + married + cit_spouse + nonfluent + spanish_hispanic_latino + central_latino + bpl_asia + medicaid + household_size + poverty + asian + black + white + other_race + employed + years_us + yrsed, train_logistic,
               method = "glm",
               trControl = control,
               metric = 'ROC')

p_logistic <- predict(logistic_model, test_logistic)


# Generate confusion matrix
logistic_matrix <- confusionMatrix(p_logistic, test_logistic$undocu_likely, positive = "X1")
print(logistic_matrix)
summary_logistic<-summary(logistic_model)$coefficients[,c(1,4)]
summary_logistic
xtable(summary_logistic, digits=4)


## Metrics for Figure 2 comparison
# Make predictions on test set
logistic.preds = predict(logistic_model, newdata = test_logistic, type = "prob")[, 2]  # Get probabilities for class "1"
# Create prediction object for ROCR
logistic.prediction = prediction(logistic.preds, test_logistic$undocu_likely)

logistic.pr = performance(logistic.prediction,"prec","rec") # Precision-Recall curve
```

# KNN

```{r KNN}
library(class)
library(caTools)

set.seed(1)
levels(all_data$knn$undocu_likely) <- make.names(levels(all_data$knn$undocu_likely))
train_index_knn <- createDataPartition(all_data$knn$undocu_likely, p = 0.7, list = FALSE)
train_knn <- all_data$knn[train_index_knn, ]
test_knn <- all_data$knn[-train_index_knn, ]



knn_model <- train(undocu_likely ~., data = train_knn, method = "knn", 
                       trControl = control, 
                       tuneLength = 10,
                       metric = 'ROC',
                       preProcess = c('center', 'scale'))

knn_model

predict_knn <- predict(knn_model, test_knn)

knn_matrix <- confusionMatrix(predict_knn, test_knn$undocu_likely, positive = "X1")
print(knn_matrix)


## Metrics for Figure 2 comparison
# Make predictions on test set
knn.preds = predict(knn_model, newdata = test_knn, type = "prob")[, 2]  # Get probabilities for class "1"
# Create prediction object for ROCR
knn.prediction = prediction(knn.preds, test_knn$undocu_likely)

knn.pr = performance(knn.prediction,"prec","rec") # Precision-Recall curve
```

-   The number of trees in the forest

-   The number of features to consider at any given split: $m_{try}$

-   The complexity of each tree

-   The sampling scheme

-   The splitting rule to use during tree construction

-   and (2) typically have the largest impact on predictive accuracy and
    should always be tuned. (3) and (4) tend to have marginal impact on
    predictive accuracy but are still worth exploring. They also have
    the ability to influence computational efficiency. (5) tends to have
    the smallest impact on predictive accuracy and is used primarily to
    increase computational efficiency.

# RF

```{r Random Forest}
library(class)
library(caTools)
library(caret)
library(rpart)  ## recursive partitioning

levels(all_data$sample$undocu_likely) <- make.names(levels(all_data$sample$undocu_likely))


train_index_rf <- createDataPartition(all_data$sample$undocu_likely, p = 0.7, list = FALSE)
train_rf <- all_data$sample[train_index_rf, ]
test_rf <- all_data$sample[-train_index_rf, ]


train_rf <- train_rf %>%
  select(-undocu_likely, undocu_likely)

#Manual search by create 10 folds and repeat 3 times
#control_rf <- trainControl(method = 'repeatedcv',
                        #number = 10,
                       # repeats = 3,
                        #search = 'grid')

tunegrid <- expand.grid(mtry = c(2,4,8,12),
                      splitrule = c("gini", "extratrees"),
                      min.node.size = 1)

control_up <- trainControl(
    method = "cv",
    number = 10,  
    summaryFunction = twoClassSummary,
    classProbs = TRUE,
    sampling = "up",
)


set.seed(1)
rf_model <- train(undocu_likely ~ age + fem + married + cit_spouse + medicaid + nonfluent + spanish_hispanic_latino + central_latino + bpl_asia + household_size + poverty + asian + black + white + other_race + employed + years_us + yrsed,
               data = train_rf,
               method = "ranger",
               trControl = control_up,
               tuneLength = 5,
               importance = "impurity",
               metric = 'ROC')
print(rf_model)
plot(rf_model)

p_rf <- predict(rf_model, test_rf)

rf_matrix <- confusionMatrix(p_rf, test_rf$undocu_likely, positive="X1")
rf_matrix


library(vip)
library(gridExtra)

feature_importance <- vip(rf_model, num_features = 19, bar = FALSE)
grid.arrange(feature_importance, nrow = 1)

all_data$dTable$undocu_logit <- predict(logistic_model, all_data$sample)
all_data$dTable$undocu_knn <- predict(knn_model, all_data$knn)
all_data$dTable$undocu_rf <- predict(rf_model, all_data$sample)
setwd("G:/Shared drives/Undocu Research/Data")
write.csv(all_data$dTable, "SIPP_dTable.csv", row.names = FALSE)


rf_model$finalModel$num.trees

# Access specific parameters
rf_optimal_trees <- rf_model$bestTune$mtry
cat("Optimal mtry:", rf_optimal_trees, "\n")

## Metrics for Figure 2 comparison
# Make predictions on test set
rf.preds = predict(rf_model, newdata = test_rf, type = "prob", n.trees = rf_optimal_trees)[, 2]  # Get probabilities for class "1"
# Create prediction object for ROCR
rf.prediction = prediction(rf.preds, test_rf$undocu_likely)

rf.pr = performance(rf.prediction,"prec","rec") # Precision-Recall curve
```

all_data$sample_num <- all_data$sample
all_data$sample_num$undocu_likely_num \<-
as.numeric(all_data$sample_num$undocu_likely) - 1

train_index_gbm \<-
createDataPartition(all_data$sample_num$undocu_likely_num, p = 0.7, list
= FALSE) train_gbm \<- all_data$sample_num[train_index_gbm, ]
test_gbm <- all_data$sample_num[-train_index_gbm, ]

gbm_model \<- gbm( undocu_likely_num \~ age + fem + married +
cit_spouse + medicaid + nonfluent + spanish_hispanic_latino +
central_latino + bpl_asia + household_size + poverty + asian + black +
white + other_race + employed + years_us + yrsed, data = train_gbm,
verbose = FALSE, cv.folds = 10, n.trees = 1766, shrinkage = 0.005,
interaction.depth = 6)

optimal_trees \<- gbm.perf(gbm_model, method = "cv", plot.it = FALSE)

# GBM

```{r Gradient-Boosting}
set.seed(1)
## Data partitioning for gbm library
train_gbm$undocu_likely <- as.numeric(train_gbm$undocu_likely) - 1      
test_gbm$undocu_likely <- as.numeric(test_gbm$undocu_likely) - 1 

gbm_model <- gbm(undocu_likely ~ age + fem + married + cit_spouse + medicaid + nonfluent + spanish_hispanic_latino + central_latino + bpl_asia + household_size + poverty + asian + black + white + other_race + employed + ln_adj + years_us + yrsed, data = train_gbm, 
                 verbose = FALSE,
                 cv.folds = 10,
                 n.trees = 713,
                 shrinkage = 0.005,
                 interaction.depth = 6)


## Model summary and optimal number of trees
gbm_model

optimal_trees_gbm <- gbm.perf(gbm_model, method = "cv", plot.it = FALSE)
cat("Optimal number of trees:", optimal_trees_gbm, "\n")

setwd(figures_path)
## Feature importance
feature_importance_gbm <- vip(gbm_model, num_features = 19, bar = FALSE)
ggsave("gbm_feature_importance.png", feature_importance_gbm, width = 22, height = 16, units = "cm", dpi = 300)
grid.arrange(feature_importance_gbm, nrow = 1)

### Save data sets and model for future use ###
saveRDS(gbm_model, file = "gbm_model.rds")
```

# College GBM

```{r}
set.seed(1)
## Data partitioning for gbm library
all_data_college$sample$undocu_likely <- as.numeric(all_data_college$sample$undocu_likely) - 1

train_index_college <- createDataPartition(all_data_college$sample$undocu_likely, p = 0.7, list = FALSE)
train_college <- all_data_college$sample[train_index_college, ]
test_college <- all_data_college$sample[-train_index_college, ]

college_gbm_model <- gbm(undocu_likely ~ age + fem + married + cit_spouse + medicaid + nonfluent + spanish_hispanic_latino + central_latino + bpl_asia + household_size + poverty + asian + black + white + other_race + employed + ln_adj + years_us + yrsed, data = train_college, 
                 verbose = FALSE,
                 cv.folds = 10,
                 n.trees = 549,
                 shrinkage = 0.005, 
                 interaction.depth = 6)


## Model summary and optimal number of trees
college_gbm_model

optimal_trees_college <- gbm.perf(college_gbm_model, method = "cv", plot.it = FALSE)
cat("Optimal number of trees:", optimal_trees_college, "\n")


setwd(figures_path)
## Feature importance
college_gbm_feature_importance <- vip(college_gbm_model, num_features = 19, bar = FALSE)
ggsave("college_gbm_feature_importance.png", college_gbm_feature_importance, width = 22, height = 16, units = "cm", dpi = 300)
grid.arrange(college_gbm_feature_importance, nrow = 1)

### Save model for future use ###
saveRDS(college_gbm_model, file = "college_gbm_model.rds")
```

# Probability Quartiles

```{r}
library(xtable)
test_gbm$p = predict(gbm_model, newdata = test_gbm, type = "response", n.trees = optimal_trees_gbm)
test_college$p = predict(gbm_model, newdata = test_college, type = "response", n.trees = optimal_trees_gbm)


# Create quartiles based on probability magnitude
test_gbm$p_quartiles <- cut(test_gbm$p, 
                                   breaks = quantile(test_gbm$p, probs = c(0, 0.25, 0.5, 0.75, 1)), 
                                   labels = c("Q1", "Q2", "Q3", "Q4"), 
                                   include.lowest = TRUE)
test_college$p_quartiles <- cut(test_college$p, 
                                   breaks = quantile(test_college$p, probs = c(0, 0.25, 0.5, 0.75, 1)), 
                                   labels = c("Q1", "Q2", "Q3", "Q4"), 
                                   include.lowest = TRUE)
# Check the distributions
table(test_gbm$p_quartiles)
table(test_college$p_quartiles)

# Proportion of undocu_likely in each quartiles
quartiles_table_gbm <- test_gbm %>%
  group_by(p_quartiles) %>%
  summarise(
    count = sum(undocu_likely == 1),
    n = n(),
    actual_proportion = (count / n)
    )
quartiles_table_college <- test_college %>%
  group_by(p_quartiles) %>%
  summarise(
    count = sum(undocu_likely == 1),
    n = n(),
    actual_proportion = (count / n)
    )

xtable(quartiles_table_gbm, digits=4)
print(quartiles_table_gbm)
xtable(quartiles_table_college, digits=4)
print(quartiles_table_college)


top_75_sample_num <- test_gbm %>%
  filter(undocu_likely == 1) %>%
  arrange(desc(p)) %>%
  mutate(
    rank = row_number(),
    cumulative_p = rank / n(),
    top_75 = cumulative_p <= 0.75
  ) %>%
  filter(top_75==TRUE)
threshold_75_num <- min(top_75_sample_num$p)


undocu_test_num <- test_gbm %>%
  filter(undocu_likely == 1) %>%
  arrange(desc(p))

# Create GBM high recall (p>threshold), high probability (75% of actual), low probability (Q1)
test_gbm$predicted_class <- ifelse(test_gbm$p >= threshold_75_num, 1, 0)

## Thersholds on undocu likely sample (BUG????)
quartile_thresholds_gbm <- quantile(undocu_test_num$p, probs = c(0, 0.25, 0.5, 0.75, 1))
print(quartile_thresholds_gbm)
```

```{r GBM Performance}
## Metrics for Figure 2 comparison (gbm library model)
# Make predictions on test set
gbm.preds = predict(gbm_model, newdata = test_gbm_num, type = "response", n.trees = optimal_trees_gbm_two)
# Create prediction object for ROCR
gbm.prediction = prediction(gbm.preds, test_gbm_num$undocu_likely)

gbm.pr = performance(gbm.prediction,"prec","rec") # Precision-Recall curve
gbm.roc = performance(gbm.prediction,"tpr","fpr")   # ROC curve 
gbm.auc <- performance(gbm.prediction,"auc")@y.values[[1]] # AUC value

gbm_auc_pr <- trapz(
  as.vector(gbm.pr@x.values)[[1]][-1], 
  as.vector(gbm.pr@y.values)[[1]][-1]
)


rf.pr = performance(rf.prediction,"prec","rec") # Precision-Recall curve
rf.roc = performance(rf.prediction,"tpr","fpr") 


setwd("G:/Shared drives/Undocu Research/Output/Figures")
png("roc_curves.png", width = 22*100, height = 16*100, res = 300)

plot(gbm.roc, main = "ROC Curves for contending models", col = "blue", lwd = 2) 
lines(rf.roc@x.values[[1]], rf.roc@y.values[[1]], col = "red", lwd = 2)

legend(x = 0.6, y = 1, 
       col = c("white", "blue", "red"),
       lwd = c(3, 3, 3), 
       lty = c(0, 1, 2),
       legend = c("Model:", "Gradient boosting", "Random Forest"),
       bty = 'n', cex = 0.95, ncol = 1, seg.len = 6)

abline(a = 0, b = 1, lty = 2, col = "gray")

dev.off()

# Plot the first curve
plot(rf.pr, col = "blue", lwd = 2, main = "Precision-Recall Curve")
# Add the second curve
lines(gbm.pr@x.values[[1]], gbm.pr@y.values[[1]], col = "red", lwd = 2)
```

# Figure 2

```{r Figure 2}
setwd("G:/Shared drives/Undocu Research/Output/Figures")
# ==============================================================================
# PREPARE PRECISION-RECALL DATA FOR ALL MODELS
# ==============================================================================

# Start with boosted trees as baseline
gbm_pr_data <- data.frame(
  recall = unlist(gbm.pr@x.values), 
  precision = unlist(gbm.pr@y.values)
)

# Interpolate other models to match boosted trees recall values
rf_pr_data <- data.frame(
  recall = approx(unlist(rf.pr@x.values), unlist(rf.pr@y.values), 
                 xout = gbm_pr_data$recall)$x,
  precision = approx(unlist(rf.pr@x.values), unlist(rf.pr@y.values), 
                    xout = gbm_pr_data$recall)$y
)

logistic_pr_data <- data.frame(
  recall = approx(unlist(logistic.pr@x.values), unlist(logistic.pr@y.values), 
                 xout = gbm_pr_data$recall)$x,
  precision = approx(unlist(logistic.pr@x.values), unlist(logistic.pr@y.values), 
                    xout = gbm_pr_data$recall)$y
)

knn_pr_data <- data.frame(
  recall = approx(unlist(knn.pr@x.values), unlist(knn.pr@y.values), 
                 xout = gbm_pr_data$recall)$x,
  precision = approx(unlist(knn.pr@x.values), unlist(knn.pr@y.values), 
                    xout = gbm_pr_data$recall)$y
)

# ==============================================================================
# COMBINE ALL DATA
# ==============================================================================

all_pr_data <- gbm_pr_data

# Add precision values for other models
all_pr_data$gbm_precision <- gbm_pr_data$precision
all_pr_data$rf_precision <- rf_pr_data$precision
all_pr_data$logistic_precision <- logistic_pr_data$precision
all_pr_data$knn_precision <- knn_pr_data$precision

# Calculate precision differences relative to boosted trees
all_pr_data$gbm_precision_difference <- gbm_pr_data$precision - all_pr_data$precision
all_pr_data$rf_precision_difference <- rf_pr_data$precision - all_pr_data$precision
all_pr_data$logistic_precision_difference <- logistic_pr_data$precision - all_pr_data$precision
all_pr_data$knn_precision_difference <- knn_pr_data$precision - all_pr_data$precision

# ==============================================================================
# PLOT 1: PRECISION-RECALL CURVES FOR ALL MODELS
# ==============================================================================

#pdf("undocumented_pr_curves.pdf", family="Times", width=9)
png("undocumented_pr_curves.png", width = 22*100, height = 16*100, res = 300, family = "Times")

op <- par(family = "serif")

plot(all_pr_data$recall, all_pr_data$precision, type = "l", pch = 19,
     col = "yellow", xlab = "Recall", ylab = "Precision",
     lwd = 3, ylim=c(0, 1), xlim = c(0,1), lty = 1, cex.lab=1.5, cex.axis=1)

lines(all_pr_data$recall, all_pr_data$rf_precision, type = "l", 
      col = "blue", lwd = 3, lty = 2)
lines(all_pr_data$recall, all_pr_data$logistic_precision, type = "l",
      col = "red", lwd = 3, lty = 3)
lines(all_pr_data$recall, all_pr_data$knn_precision, type = "l",
      col = "green", lwd = 3, lty = 4)

legend(x = 0.5, y = 1, 
       col = c("white", "yellow", "blue", "red", "green"),
       lwd = c(3, 3, 3, 3, 3, 3), 
       lty = c(0, 1, 2, 3, 4, 5),
       legend = c("Model:", "Gradient-Boosting Trees", "Random Forest", "Logistic", "KNN"),
       bty = 'n', cex = 1.25, ncol = 1, seg.len = 6)

# Add random baseline
total_undocu <- sum(test_gbm_num$undocu_likely == 1)
random_baseline <- total_undocu / nrow(test_gbm_num)
abline(h = random_baseline, lty = 2, col = "gray")
# Add text label for random baseline
text(0.5, random_baseline + 0.05, 
     paste("Random baseline:", round(random_baseline, 3)), 
     cex = 0.9, col = "gray30")


dev.off()


### Relative Pr-Re curves ###

png("undocumented_precision_differences.png", width = 22*100, height = 16*100, res = 300, family = "Times")

op <- par(family = "serif")

plot(all_pr_data$recall, all_pr_data$rf_precision_difference, type = "l", pch = 19, 
     col = "blue", xlab = "Recall Relative to Boosted Trees", ylab = "Precision", 
     lwd = 3, ylim = c(-0.3, 0.1), xlim = c(0,1), lty = 2, cex.lab=1.5, cex.axis=1)

lines(all_pr_data$recall, all_pr_data$logistic_precision_difference, type = "l", 
      col = "red", lwd = 3, lty = 3)
lines(all_pr_data$recall, all_pr_data$knn_precision_difference, type = "l",
      col = "green", lwd = 3, lty = 4)

legend(x = 0.5, y = -0.1, 
       col = c("white", "blue", "red", "green"),
       lwd = c(3, 3, 3, 3), 
       lty = c(0, 2, 3, 4),
       legend = c("Model:", "Random Forest", "Logistic", "KNN"),
       bty = 'n', cex = 1.25, ncol = 1, seg.len = 5)

abline(h = 0, lty = 2, col = "gray")
# Add text label for random baseline
text(0.5, 0.05, 
     paste("GBM baseline:", round(0, 3)), 
     cex = 0.9, col = "gray30")

dev.off()
```

*undocu_likely* age *married* nonfluent *spanish_hispanic_latino*
central_latino *bpl_asia* household_size *poverty* asian *black* white
*other_race* employed *years_us* yrsed

<https://bradleyboehmke.github.io/HOML/process.html> Sensitivity:
$\frac{TP}{TP + FN}$ Specificity: $\frac{TN}{TN + FP}$ Precision /
Positive-predictive value: $\frac{TP}{TP + FP}$ Accuracy:
$\frac{TP + TN}{total}$

# ACS

```{r Loading ACS data}
library(readr)
setwd(data_path)

ACS_variables <- c("hisp", "spanish", "race", "bpld", "empstat", "poverty", "married", "cit_spouse", "nonfluent", "bpl_asia", "asian", "black", "white", "fem", "numprec", "yrsusa1", "hinscare", "hinscaid", "undocu", "bpl_usa", "age", "yrsed")  


ACS <- read_csv("ML Training Sample.csv")
View(ACS)
```

```{r Preparing ACS data}
ACS_sipp <- ACS %>%
  mutate(
    spanish_hispanic_latino = as.factor(ifelse(hisp==1 | spanish==1, 1, 0)),
    central_latino = as.factor(ifelse((hisp==1) & (bpld=="belize/british honduras" | bpld=="costa rica" | bpld=="el salvador" | bpld=="guatemala" | bpld=="honduras" | bpld=="nicaragua" | bpld=="panama" | bpld=="mexico"), 1, 0)),
    black = as.factor(ifelse(race=="black/african america", 1, 0)),
    white = as.factor(ifelse(race=="white", 1, 0)),
    employed = as.factor(ifelse(empstat==1, 1, 0)),
    poverty = as.factor(ifelse(poverty<100, 1, 0)),
    married = as.factor(married),
    cit_spouse = as.factor(cit_spouse),
    nonfluent = as.factor(nonfluent),
    bpl_asia = as.factor(bpl_asia),
    asian = as.factor(asian),
    fem = as.factor(fem),
    household_size = numprec,
    #Change SIPP 0 years_us to NA
    years_us = yrsusa1,
    medicare = as.factor(ifelse(hinscare=="yes", 1, 0)),
    medicaid = as.factor(ifelse(hinscaid=="has insurance through medicaid", 1, 0))
    
  )
ACS_sipp$years_us <- replace(ACS_sipp$years_us, ACS_sipp$yrsusa1=="n/a or less than one year" & ACS_sipp$bpl_usa==1, NA)
ACS_sipp$years_us <- replace(ACS_sipp$years_us, ACS_sipp$yrsusa1=="n/a or less than one year" & ACS_sipp$bpl_usa==0, 0)
ACS_sipp$years_us <- as.numeric(ACS_sipp$years_us)

ACS_sipp$other_race <- as.factor(ifelse(ACS$black!=1 & ACS$white!=1 & ACS$asian!=1, 1, 0))
ACS_sipp_na <- ACS_sipp  %>%
  filter(undocu==1, !is.na(years_us),!is.na(medicaid), !is.na(age),!is.na(fem), !is.na(married),!is.na(cit_spouse), !is.na(nonfluent),  !is.na(spanish_hispanic_latino), !is.na(central_latino), !is.na(bpl_asia), !is.na(household_size), !is.na(poverty), !is.na(asian), !is.na(black), !is.na(white), !is.na(other_race), !is.na(employed),  !is.na(yrsed))

ACS_cols_numeric <- c("central_latino", "bpl_asia", "medicaid", "age", "fem", "married", "cit_spouse", "nonfluent", "spanish_hispanic_latino", "household_size", "poverty", "asian", "black", "white", "other_race", "employed", "years_us","yrsed")

ACS_sipp_knn <- ACS_sipp_na %>%
  mutate_at(ACS_cols_numeric, as.numeric)
```

```{r Undocu imputation}
setwd(data_path)
## Imputation method 1: Logistic regression 
ACS_sipp_na$undocu_logistic <- predict(logistic_model, ACS_sipp_na)
write.csv(ACS_sipp_na, "ACS_SIPP_logistic.csv", row.names = FALSE)

## Imputation method 2: KNN 
ACS_sipp_na$knn_undocu <- predict(knn_model, ACS_sipp_knn)
write.csv(ACS_sipp_na, "ACS_SIPP_knn.csv", row.names = FALSE)

## Imputation method 3: RF 
ACS_sipp_na$rf_undocu <- predict(rf_model, ACS_sipp_na)
write.csv(ACS_sipp_na, "ACS_SIPP_rf.csv", row.names = FALSE)

## Imputation method 4: GBM
setwd(data_path)

ACS_sipp_na <- ACS_sipp_na %>%
  mutate(
    gbm_undocu_p = predict(gbm_model, type = "response", ACS_sipp_na, n.trees = optimal_trees_gbm_two),
    
    gbm_undocu_q = case_when(
      gbm_undocu_p < quartile_thresholds_gbm[2] ~ "Q1",
      gbm_undocu_p > quartile_thresholds_gbm[2] & gbm_undocu_p < quartile_thresholds_gbm[3] ~ "Q2",
      gbm_undocu_p > quartile_thresholds_gbm[3] & gbm_undocu_p < quartile_thresholds_gbm[4] ~ "Q3",
      gbm_undocu_p > quartile_thresholds_gbm[4] ~ "Q4",
      TRUE ~ "Unknown"
      ),
    gbm_high_prob = ifelse(gbm_undocu_q=="Q4", 1, 0),
    gbm_low_prob = ifelse(gbm_undocu_q=="Q1", 1, 0),
    gbm_high_recall = ifelse(gbm_undocu_p > threshold_75_num, 1, 0)
    
    
    ### College P quartiles and Thresholds
  )

write.csv(ACS_sipp_na, "ACS_SIPP_gbm.csv", row.names = FALSE)
```

# ML Metrics

```{r}
# Make predictions on test set
boost.pred = predict(caret_gbm_model, newdata = test_gbm, type = "prob", n.trees = optimal_trees_gbm)[, 2]  # Get probabilities for class "1"
# Create prediction object for ROCR
boost.prediction = prediction(boost.pred, test_gbm$undocu_likely)
# Get raw predictions (class labels)
boost.predict = predict(caret_gbm_model, newdata = test_gbm, n.trees=optimal_trees_gbm)  # Class predictions

boost.roc = performance(boost.prediction,"tpr","fpr")   # ROC curve 
boost.pr = performance(boost.prediction,"prec","rec") # Precision-Recall curve
boost.auc <- performance(boost.prediction,"auc")@y.values[[1]] # AUC value

boost_auc_pr <- trapz(
  as.vector(boost.pr@x.values)[[1]][-1], 
  as.vector(boost.pr@y.values)[[1]][-1]
)

gbm_performance <- boost.pr
plot(boost.roc, main = "ROC Curve", col = "blue", lwd = 2) #PLOT EACH ON A SINGLE GRAPH FOR ROC
abline(a = 0, b = 1, lty = 2, col = "red")
plot(boost.pr) # plot Precision-Recall curve


cat("Optimal number of trees:", optimal_trees_gbm, "\n")
cat("AUC-ROC:", round(boost.auc, 4), "\n")
cat("AUC-PR:", round(boost_auc_pr, 4), "\n")

predicted_class <- ifelse(boost.pred > 0.5, 1, 0)
table(Predicted = predicted_class, Actual = test_gbm$undocu_likely)
plot(caret_gbm_model)

# Feature Importance
library(vip)
library(gridExtra)

feature_importance_gbm <- vip(caret_gbm_model, num_features = 19, bar = FALSE)
grid.arrange(feature_importance_gbm, nrow = 1)
```
