---
title: "Minimal SIPP-to-ACS GBM Prediction Export"
format: html
---

## 1. Setup

```{r}
library(dplyr)
library(readr)
library(xgboost)
library(caret)

if (Sys.info()["sysname"] == "Windows") {
  base_dir <- "G:/Shared drives/Undocu Research"
} else {
  base_dir <- "~/Library/CloudStorage/GoogleDrive-vsovero@ucr.edu/Shared drives/Undocu Research"
}

data_path <- file.path(base_dir, "Data")
```

## 2. Read cleaned SIPP and ACS files

```{r}
sipp <- read_csv(
  file.path(data_path, "(Step 1 output) Core_TM SIPP 2008 Wave 2.csv"),
  show_col_types = FALSE
)

acs <- read_csv(
  file.path(data_path, "ACS Target Sample.csv"),
  show_col_types = FALSE
)
```

## 3. Define features and prepare model data

```{r}
# Same model features as the working target-like comparison file.
# Race enters as harmonized race dummies.
features <- c(
  "age", "fem", "married", "cit_spouse", "medicaid", "nonfluent",
  "spanish_hispanic_latino", "central_latino", "bpl_asia",
  "household_size", "poverty", "asian", "black", "white", "other_race",
  "employed", "years_us", "years_us_missing", "yrsed"
)


# Explicit age training-sample restriction.
# This avoids relying on missing yrsed to implicitly drop children outside
# the education-question universe.
sipp<- sipp %>%
  filter(age >= 15)

# Keep rows complete on required model variables.
# After the explicit age restriction, this removes only unresolved predictor
# missingness rather than defining the age universe indirectly.

sipp <- sipp[complete.cases(sipp[, c("undocu_likely", features, "college")]), ]

# ACS needs model features plus merge keys for export back to Stata.
acs <- acs[complete.cases(acs[, c("year", "serial", "pernum", features)]), ]
```

## 4. Split SIPP into training and test samples

```{r}
set.seed(1)

train_index <- createDataPartition(sipp$undocu_likely, p = 0.70, list = FALSE)

train_sipp <- sipp[train_index, ]
test_sipp  <- sipp[-train_index, ]

train_y <- as.numeric(train_sipp$undocu_likely)
test_y <- as.numeric(test_sipp$undocu_likely)

train_matrix <- train_sipp %>%
  select(all_of(features)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

test_matrix <- test_sipp %>%
  select(all_of(features)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

dtrain <- xgb.DMatrix(data = train_matrix, label = train_y)
dtest <- xgb.DMatrix(data = test_matrix, label = test_y)
```

## 5. Train XGBoost model using original grid-search logic

```{r}
xgb_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1),
  max_depth = c(4, 6, 8),
  subsample = c(0.8, 1.0),
  colsample_bytree = c(0.8, 1.0)
)

xgb_grid$optimal_trees <- 0
xgb_grid$max_cv_auc <- 0

set.seed(1)

for(i in 1:nrow(xgb_grid)) {

  params <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    eta = xgb_grid$eta[i],
    max_depth = xgb_grid$max_depth[i],
    subsample = xgb_grid$subsample[i],
    colsample_bytree = xgb_grid$colsample_bytree[i]
  )

  cv_results <- xgb.cv(
    params = params,
    data = dtrain,
    nrounds = 5000,
    nfold = 5,
    early_stopping_rounds = 50,
    maximize = TRUE,
    verbose = 0
  )

  eval_log <- cv_results$evaluation_log

  xgb_grid$max_cv_auc[i] <- max(eval_log$test_auc_mean)
  xgb_grid$optimal_trees[i] <- which.max(eval_log$test_auc_mean)
}

best_xgb_params <- xgb_grid[which.max(xgb_grid$max_cv_auc), ]
print(best_xgb_params)

final_xgb_params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  eta = best_xgb_params$eta,
  max_depth = best_xgb_params$max_depth,
  subsample = best_xgb_params$subsample,
  colsample_bytree = best_xgb_params$colsample_bytree
)

set.seed(1)

xgb_model <- xgb.train(
  params = final_xgb_params,
  data = dtrain,
  nrounds = best_xgb_params$optimal_trees,
  verbose = 0
)
```

## 6. Predict SIPP test sample and ACS target sample

```{r}
# Use GBM naming convention for downstream Stata compatibility.
test_sipp$gbm_undocu_p <- predict(xgb_model, newdata = dtest)

acs_matrix <- acs %>%
  select(all_of(features)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

dacs <- xgb.DMatrix(data = acs_matrix)

acs$gbm_undocu_p <- predict(xgb_model, newdata = dacs)
```

## 7. Define SIPP target-like reference group for quartiles

```{r}
# Reference group for quartile cutpoints:
# SIPP test observations that mirror the ACS target sample:
# college graduates, employed, ages 22-55.
sipp_targetlike_test <- test_sipp %>%
  filter(
    college == 1,
    employed == 1,
    age >= 22,
    age <= 55
  )
```

## 8. Create quartiles from SIPP target-like test probabilities

```{r}
# Quartile thresholds are based on SIPP target-like test probabilities.
# ACS is then classified using these same thresholds.
sipp_breaks <- quantile(
  sipp_targetlike_test$gbm_undocu_p,
  probs = c(0, .25, .50, .75, 1),
  na.rm = TRUE
)

# Extend endpoints so ACS observations outside the SIPP target-like range
# are retained rather than coded as NA by cut().
sipp_breaks[1] <- -Inf
sipp_breaks[length(sipp_breaks)] <- Inf

sipp_targetlike_test$gbm_undocu_q <- cut(
  sipp_targetlike_test$gbm_undocu_p,
  breaks = sipp_breaks,
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4")
)

acs$gbm_undocu_q <- cut(
  acs$gbm_undocu_p,
  breaks = sipp_breaks,
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4")
)
```

## 9. Define high-recall threshold using original full-test-positive logic

```{r}
# Original high-recall definition:
# among true positives in the full SIPP test sample, sort by predicted probability
# and choose the minimum probability among the top 75% of positives.
top_75_sample_gbm <- test_sipp %>%
  filter(undocu_likely == 1) %>%
  arrange(desc(gbm_undocu_p)) %>%
  mutate(
    rank = row_number(),
    cumulative_p = rank / n(),
    top_75 = cumulative_p <= 0.75
  ) %>%
  filter(top_75 == TRUE)

threshold_75_gbm <- min(top_75_sample_gbm$gbm_undocu_p)

threshold_75_gbm
```

## 10. Check quartile distributions

```{r}
sipp_quartiles <- sipp_targetlike_test %>%
  group_by(gbm_undocu_q) %>%
  summarise(
    n = n(),
    undocu_likely_n = sum(undocu_likely == 1),
    undocu_likely_share = mean(undocu_likely == 1),
    mean_p = mean(gbm_undocu_p),
    .groups = "drop"
  )

acs_quartiles <- acs %>%
  group_by(gbm_undocu_q) %>%
  summarise(
    n = n(),
    share = n() / nrow(acs),
    mean_p = mean(gbm_undocu_p),
    .groups = "drop"
  )

sipp_quartiles
acs_quartiles
```

## 11. Export ACS predictions for Stata

```{r}
acs_gbm_export <- acs %>%
  mutate(
    gbm_high_prob = ifelse(gbm_undocu_q == "Q4", 1, 0),
    gbm_low_prob = ifelse(gbm_undocu_q == "Q1", 1, 0),
    gbm_high_recall = ifelse(gbm_undocu_p >= threshold_75_gbm, 1, 0)
  ) %>%
  select(
    gbm_undocu_p,
    gbm_undocu_q,
    gbm_high_prob,
    gbm_low_prob,
    gbm_high_recall,
    year,
    serial,
    pernum
  )

write_csv(
  acs_gbm_export,
  file.path(data_path, "ACS_SIPP_gbm.csv")
)
```
